Global code motion of congruent instructions

### 2017-04-17 paper.tex
tables

Dhamdhere \cite{dhamdhere1988fast}, Muchnick \cite{steven1997advanced} mention
code hoisting in a data flow framework. A list of Very Busy Expressions (VBE)
computed which are hoisted in a basic block where the expression is
anticipable (all the operands are available). This algorithm would hoist as far
as possible without regarding the impact on register pressure and as such a cost
model will be required.  Also the description of VBE is based on the classic
dataflow model and an adaptation to a sparse \SSA{} representation is required.

Rosen \cite{rosen1988global} also briefly discuss hoisting computations with
identical value numbers from immediate successors. Their algorithm iterates on
computations of same rank and move the code with identical computations from the
sibling branch to a common dominator if they are very busy
\cite{steven1997advanced}. The cost-model to mitigate register pressure is
missing, also there is no mention of sinking congruent instructions.

@@

Click \cite{click1995global} describe aggressive global code motion to first
schedule all the instructions as early as possible. This results in very long
live ranges which is mitigated by again scheduling all the instructions as late
as possible. They report sppeedup of as high as $23\%$.

Barany \cite{barany2013} presented a global scheduler with ILP formulation with
a goal to minimize register pressure. The results they got were not very
promising. It may be because they only used the scheduler for smaller functions
($<1000$ instructions); also, they compiled the benchmarks for ARM-Cortex which
is more resilient to register pressure because it has more registers compared to
X86, for example.

Shobaki \cite{shobaki2013} also recently presented a combinatorial global
scheduler with reasonable performance improvements. It is possible that both
Barany and Shobaki's implmentation will have similar results when compiled for
same target architecture. Also, both suffer from the same problem, although
Shobaki not so much, of large compile times which is not feasible for industrial
compilers like \GCC{} and \LLVM{}.

@@

The amount of hoisting depends on whether we collect \GVN{} of instructions
before finding candidates (optimistic) or, on-demand (pessimistic). It also
depends on the generality of the \GVN{} algorithm as mentioned earlier in
Section~\ref{subsec:finding-candidates}. We have implemented an optimistic
global code motion of congruent instructions which uses the liveness analysis as
illustrated in Das \cite{das2012} and ranking expressions explained in Rosen
\cite{rosen1988global}.

@@

Code size also remained almost the same as a result of code-motion as shown in
Table~\cite{tab:code-size}.  The code size listed here is the text size (from
linux size command) of the final executables for each benchmark.  Some bencmarks
like 473.astar increased in code size by almost 2\%. This is because once the
code-size of a function decreases it becomes cheaper to inline. As we can see in
Table-\ref{tab:stats}, four more functions got inlined in 473.astar. Except for
447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk all other benchmarks got
more functions inlined.

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|r|r|r|}
      \hline
Number of spills  &base(-Ofast) &code-motion(-Ofast) &\%change \\\hline
400.perlbench	  & 2539	& 2493	& 0.98 \\\hline
401.bzip2	  & 718	        & 707	& 0.98 \\\hline
403.gcc	          & 5782	& 5722	& 0.99 \\\hline
429.mcf	          & 14	        & 17	& 1.21 \\\hline
433.milc	  & 635	        & 642	& 1.01 \\\hline
444.namd	  & 3223	& 3274	& 1.01 \\\hline
445.gobmk	  & 2166	& 2173	& 1.00 \\\hline
447.dealII	  & 11074	& 10234	& 0.92 \\\hline
450.soplex	  & 1122	& 1125	& 1.00 \\\hline
453.povray	  & 4701	& 4692	& 1.09 \\\hline
456.hmmer	  & 1197	& 1274	& 1.06 \\\hline
458.sjeng	  & 168	        & 176	& 1.05 \\\hline
462.libquantum	  & 118	        & 118	& 1.00 \\\hline
464.h264ref	  & 3379	& 3454	& 1.02 \\\hline
470.lbm	          & 33	        & 33	& 1.00 \\\hline
471.omnetpp	  & 524	        & 527	& 1.00 \\\hline
473.astar	  & 180	        & 201	& 1.12 \\\hline
482.sphinx3	  & 674	        & 670	& 0.99 \\\hline
483.xalancbmk	  & 5102	& 4965	& 0.97 \\\hline
Grand Total	  & 43349	& 42497	& 0.98 \\\hline
    \end{tabular}
  \end{center}
  \caption{Number of spills on before and after code-motion on SPEC2006 at -Ofast}
  \label{tab:spills}
\end{table}

Spills are listed in Table~\cite{tab:spills}. Spills were reduced by an average of 2\% on the
\SPEC{} benchmark suite. This is a result of cost-model which limits the code-motion to those
candidates which do not increase the register pressure (except for loads Section~\cite{sec:cost-models},
overall spills still decreases.

\begin{table}[h]
  \begin{center}
    \begin{tabular}{|l|c|c|c|}
      \hline
  Benchmarks      &   Baseline	        &  GCM & \%Increase     \\\hline
  400.perlbench	  &   186,829,426	& 184,690,006	& 0.99 \\\hline
  401.bzip2	  &   76,337,033	& 75,942,641	& 0.99 \\\hline
  403.gcc         &   1,079,625,231	& 1,073,158,220	& 0.99 \\\hline
  429.mcf         &   78,336,298	& 77,859,513	& 0.99 \\\hline
  433.milc	  &   272,525,306	& 269,684,876	& 0.99 \\\hline
  444.namd	  &   78,354,313	& 77,875,603	& 0.99 \\\hline
  445.gobmk	  &   261,229,078	& 258,459,798	& 0.99 \\\hline
  447.dealII	  &   643,457,797	& 638,628,331	& 0.99 \\\hline
  450.soplex	  &   245,442,755	& 242,811,500	& 0.99 \\\hline
  453.povray	  &   499,034,715	& 494,785,035	& 0.99 \\\hline
  456.hmmer	  &   214,425,324	& 211,984,389	& 0.99 \\\hline
  458.sjeng	  &   89,612,524	& 88,803,166	& 0.99 \\\hline
  462.libquantum  &   84,762,158	& 84,077,905	& 0.99 \\\hline
  464.h264ref	  &   151,049,588	& 149,284,624	& 0.99 \\\hline
  470.lbm         &   71,878,726	& 71,775,405	& 1.00 \\\hline
  471.omnetpp	  &   382,661,268	& 379,123,833	& 0.99 \\\hline
  473.astar	  &   78,351,117	& 77,872,820	& 0.99 \\\hline
  482.sphinx3	  &   158,434,290	& 156,586,406	& 0.99 \\\hline
  483.xalancbmk	  &   22,485,708,759	& 22,455,060,569& 1.00 \\\hline
    \end{tabular}
  \end{center}
  \caption{Change in compilation time with and without code-motion on SPEC2006 at -Ofast}
  \label{tab:compile-time}
\end{table}

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|r|r|r|}
      \hline
  &  \multicolumn{3}{|c|}{Hoisted by LICM} &\multicolumn{3}{|c|}{Sunk by LICM} &\multicolumn{3}{|c|}{Functions Inlined} &  \multicolumn{3}{|c|}{Functions Deleted} \\\hline
 Benchmarks       & Base        & GCM	& \%  &Base     & GCM   &\%    & Base	& GCM	  &\%   & Base	& GCM	& \% \\\hline
 400.perlbench	  & 2428	& 2480	& 2.1 & 36	& 35	& -2.8 & 1562	& 1565	  & 0.2 & 232	& 234	& 0.9  \\\hline
 401.bzip2	  & 721	        & 825	& 14.4& 3	& 3	& 0.0  & 172	& 180	  & 4.7 & 32	& 33	& 3.1  \\\hline
 403.gcc	  & 8867	& 10079	& 13.7& 143	& 182	& 27.3 & 5034	& 5080	  & 0.9 & 1303	& 130	& 0.1  \\\hline
 429.mcf	  & 55	        & 51	& -7.3& 1	& 1	& 0.0  & 4	& 4	  & 0.0 & 0 	& 0	& 0.0  \\\hline
 433.milc	  & 2537	& 2795	& 10.2& 3	& 3	& 0.0  & 75	& 81	  & 8.0 & 0	& 0	& 0.0  \\\hline
 444.namd	  & 2204	& 2206	& 0.1 & 240	& 240	& 0.0  & 387	& 390	  & 0.8 & 57	& 55	& -3.5 \\\hline
 445.gobmk	  & 3090	& 3107	& 0.6 & 18	& 19	& 5.6  & 735	& 775	  & 5.4 & 230	& 234	& 1.7  \\\hline
 447.dealII	  & 59114	& 95884	& 62.2& 60	& 8	& -86.7& 100891& 100692  & -0.2& 30219	& 29966	& -0.8 \\\hline
 450.soplex	  & 4668	& 4848	& 3.9 & 18	& 17	& -5.6 & 9801	& 9784	  & -0.2& 3125	& 3112	& -0.4 \\\hline
 453.povray	  & 9533	& 9079	& -4.8& 23	& 32	& 39.1 & 5522	& 5532	  & 0.2 & 1202	& 1201	& -0.1 \\\hline
 456.hmmer	  & 3200	& 3232	& 1.0 & 3	& 8	& 166.7& 562	& 562	  & 0.0 & 103	& 101	& -1.9 \\\hline
 458.sjeng	  & 400	        & 446	& 11.5& 0	& 0	& 0.0  & 127	& 135	  & 6.3 & 18	& 18	& 0.0  \\\hline
 462.libquantum	  & 362	        & 386	& 6.6 & 0 	& 0	& 0.0  & 87	& 87	  & 0.0 & 21	& 21	& 0.0  \\\hline
 464.h264ref	  & 17075	& 17153	& 0.5 & 4	& 4	& 0.0  & 338	& 382	  & 13.0& 72	& 72	& 0.0  \\\hline
 470.lbm	  & 39	        & 39	& 0.0 & 0	& 0	& 0.0  & 13	& 13	  & 0.0 & 4	& 4	& 0.0  \\\hline
 471.omnetpp	  & 565	        & 661	& 17.0& 32	& 30	& -6.3 & 4790	& 4843	  & 1.1 & 1321	& 1316	& -0.4 \\\hline
 473.astar	  & 691	        & 754	& 9.1 & 0	& 0	& 0.0  & 359	& 363	  & 1.1 & 113	& 113	& 0.0  \\\hline
 482.sphinx3	  & 1599	& 1693	& 5.9 & 11	& 9	& -18.2& 198	& 194	  & -2.0& 53	& 53	& 0.0  \\\hline
 483.xalancbmk	  & 9997	& 11453	& 14.6& 56	& 54	& -3.6 & 69534	& 68839	  & -1.0& 29875	& 29519	& -1.2 \\\hline
    \end{tabular}
  \end{center}
  \caption{Static metrics before and after code-hoisting on SPEC2006 at -Ofast. The \% Column show percentage increase in metric w.r.t. Base. GCM improves LICM and inlining}
  \label{tab:stats}
\end{table}

Some useful statc metric are presented in Table~\cite{tab:stats}, which shows how code-motion
impacted important compiler optimizations like inlining and LICM. LICM removes loop invariant
code out of the loop. It may hoist the code to the loop's preheader or sink the code to loop's
post-dominator \cite{steven1997advanced}. As we can see number of instructions hoisted increased
significantly in several benchmarks, although number of instructions sunk did not change by much.
Number of inlined functiosn also improved for most benchmarks. In general when more functions
were inlined, more functions were deleted as in 400.perlbench, 445.gobmk.

>>>>

\citet{dhamdhere1988fast}, \citet{steven1997advanced} mention
code hoisting in a data flow framework. A list of Very Busy Expressions (VBE)
computed which are hoisted in a basic block where the expression is
anticipable (all the operands are available). This algorithm would hoist as far
as possible without regarding the impact on register pressure and as such a cost
model will be required.  Also the description of VBE is based on the classic
dataflow model and an adaptation to a sparse \SSA{} representation is required.

\citet{rosen1988global} also briefly discuss hoisting computations with
identical value numbers from immediate successors. Their algorithm iterates on
computations of same rank and move the code with identical computations from the
sibling branch to a common dominator if they are very busy
\cite{steven1997advanced}. The cost-model to mitigate register pressure is
missing, also there is no mention of sinking congruent instructions.

@@

\citet{click1995global} describes aggressive global code motion to first
schedule all the instructions as early as possible. This results in very long
live ranges which is mitigated by again scheduling all the instructions as late
as possible. They report sppeedup of as high as $23\%$.

\citet{barany2013} presented a global scheduler with ILP formulation with
a goal to minimize register pressure. The results they got were not very
promising. It may be because they only used the scheduler for smaller functions
($<1000$ instructions); also, they compiled the benchmarks for ARM-Cortex which
is more resilient to register pressure because it has more registers compared to
X86, for example.

\citet{shobaki2013} also recently presented a combinatorial global
scheduler with reasonable performance improvements. It is possible that both
Barany and Shobaki's implmentation will have similar results when compiled for
same target architecture. Also, both suffer from the same problem, although
Shobaki not so much, of large compile times which is not feasible for industrial
compilers like \GCC{} and \LLVM{}.

@@

The amount of hoisting depends on whether we collect \GVN{} of instructions
before finding candidates (optimistic) or, on-demand (pessimistic). It also
depends on the generality of the \GVN{} algorithm as mentioned earlier in
Section~\ref{subsec:finding-candidates}. We have implemented an optimistic
global code motion of congruent instructions which uses the liveness analysis as
illustrated by \citet{das2012} and ranking expressions explained by
\citet{rosen1988global}.

@@

Code size also remained almost the same as a result of code-motion as shown in
Table~\ref{tab:code-size}.  The code size listed here is the text size (from
linux size command) of the final executables for each benchmark.  Some bencmarks
like 473.astar increased in code size by almost 2\%. This is because once the
code-size of a function decreases it becomes cheaper to inline. As we can see in
Table-\ref{tab:stats}, four more functions got inlined in 473.astar. Except for
447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk all other benchmarks got
more functions inlined.

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|r|r|r|}
      \hline
Number of spills  &base(-Ofast) &GCM(-Ofast) &\%change \\\hline
400.perlbench	  & 2539	& 2493	& 0.98 \\\hline
401.bzip2	  & 718	        & 707	& 0.98 \\\hline
403.gcc	          & 5782	& 5722	& 0.99 \\\hline
429.mcf	          & 14	        & 17	& 1.21 \\\hline
433.milc	  & 635	        & 642	& 1.01 \\\hline
444.namd	  & 3223	& 3274	& 1.01 \\\hline
445.gobmk	  & 2166	& 2173	& 1.00 \\\hline
447.dealII	  & 11074	& 10234	& 0.92 \\\hline
450.soplex	  & 1122	& 1125	& 1.00 \\\hline
453.povray	  & 4701	& 4692	& 1.09 \\\hline
456.hmmer	  & 1197	& 1274	& 1.06 \\\hline
458.sjeng	  & 168	        & 176	& 1.05 \\\hline
462.libquantum	  & 118	        & 118	& 1.00 \\\hline
464.h264ref	  & 3379	& 3454	& 1.02 \\\hline
470.lbm	          & 33	        & 33	& 1.00 \\\hline
471.omnetpp	  & 524	        & 527	& 1.00 \\\hline
473.astar	  & 180	        & 201	& 1.12 \\\hline
482.sphinx3	  & 674	        & 670	& 0.99 \\\hline
483.xalancbmk	  & 5102	& 4965	& 0.97 \\\hline
Grand Total	  & 43349	& 42497	& 0.98 \\\hline
    \end{tabular}
  \end{center}
  \caption{Number of spills before and after global-code-motion (GCM) on SPEC2006 at -Ofast}
  \label{tab:spills}
\end{table}

Spills are listed in Table~\ref{tab:spills}. Spills were reduced by an average
of 2\% on the \SPEC{} benchmark suite. This is a result of cost-model which
limits the code-motion to those candidates which do not increase the register
pressure (except for loads Section~\ref{sec:cost-models}, overall spills still
decreases.

\begin{table}[h]
  \begin{center}
    \begin{tabular}{|l|c|c|c|}
      \hline
  Benchmarks      &   Baseline	        & GCM           & \%Increase \\\hline
  400.perlbench	  &   186,829,426	& 184,690,006	& 0.99 \\\hline
  401.bzip2	  &   76,337,033	& 75,942,641	& 0.99 \\\hline
  403.gcc         &   1,079,625,231	& 1,073,158,220	& 0.99 \\\hline
  429.mcf         &   78,336,298	& 77,859,513	& 0.99 \\\hline
  433.milc	  &   272,525,306	& 269,684,876	& 0.99 \\\hline
  444.namd	  &   78,354,313	& 77,875,603	& 0.99 \\\hline
  445.gobmk	  &   261,229,078	& 258,459,798	& 0.99 \\\hline
  447.dealII	  &   643,457,797	& 638,628,331	& 0.99 \\\hline
  450.soplex	  &   245,442,755	& 242,811,500	& 0.99 \\\hline
  453.povray	  &   499,034,715	& 494,785,035	& 0.99 \\\hline
  456.hmmer	  &   214,425,324	& 211,984,389	& 0.99 \\\hline
  458.sjeng	  &   89,612,524	& 88,803,166	& 0.99 \\\hline
  462.libquantum  &   84,762,158	& 84,077,905	& 0.99 \\\hline
  464.h264ref	  &   151,049,588	& 149,284,624	& 0.99 \\\hline
  470.lbm         &   71,878,726	& 71,775,405	& 1.00 \\\hline
  471.omnetpp	  &   382,661,268	& 379,123,833	& 0.99 \\\hline
  473.astar	  &   78,351,117	& 77,872,820	& 0.99 \\\hline
  482.sphinx3	  &   158,434,290	& 156,586,406	& 0.99 \\\hline
  483.xalancbmk	  &   22,485,708,759	& 22,455,060,569& 1.00 \\\hline
    \end{tabular}
  \end{center}
  \caption{Change in compilation time with and without global-code-motion (GCM) on SPEC2006 at -Ofast}
  \label{tab:compile-time}
\end{table}

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|r|r|r|}
      \hline
  &  \multicolumn{3}{|c|}{Hoisted by LICM} &\multicolumn{3}{|c|}{Sunk by LICM} &\multicolumn{3}{|c|}{Functions Inlined} &  \multicolumn{3}{|c|}{Functions Deleted} \\\hline
 Benchmarks       & Base        & GCM	& \%  &Base     & GCM   &\%    & Base	& GCM	  &\%   & Base	& GCM	& \% \\\hline
 400.perlbench	  & 2428	& 2480	& 2.1 & 36	& 35	& -2.8 & 1562	& 1565	  & 0.2 & 232	& 234	& 0.9  \\\hline
 401.bzip2	  & 721	        & 825	& 14.4& 3	& 3	& 0.0  & 172	& 180	  & 4.7 & 32	& 33	& 3.1  \\\hline
 403.gcc	  & 8867	& 10079	& 13.7& 143	& 182	& 27.3 & 5034	& 5080	  & 0.9 & 1303	& 130	& 0.1  \\\hline
 429.mcf	  & 55	        & 51	& -7.3& 1	& 1	& 0.0  & 4	& 4	  & 0.0 & 0 	& 0	& 0.0  \\\hline
 433.milc	  & 2537	& 2795	& 10.2& 3	& 3	& 0.0  & 75	& 81	  & 8.0 & 0	& 0	& 0.0  \\\hline
 444.namd	  & 2204	& 2206	& 0.1 & 240	& 240	& 0.0  & 387	& 390	  & 0.8 & 57	& 55	& -3.5 \\\hline
 445.gobmk	  & 3090	& 3107	& 0.6 & 18	& 19	& 5.6  & 735	& 775	  & 5.4 & 230	& 234	& 1.7  \\\hline
 447.dealII	  & 59114	& 95884	& 62.2& 60	& 8	& -86.7& 100891& 100692  & -0.2& 30219	& 29966	& -0.8 \\\hline
 450.soplex	  & 4668	& 4848	& 3.9 & 18	& 17	& -5.6 & 9801	& 9784	  & -0.2& 3125	& 3112	& -0.4 \\\hline
 453.povray	  & 9533	& 9079	& -4.8& 23	& 32	& 39.1 & 5522	& 5532	  & 0.2 & 1202	& 1201	& -0.1 \\\hline
 456.hmmer	  & 3200	& 3232	& 1.0 & 3	& 8	& 166.7& 562	& 562	  & 0.0 & 103	& 101	& -1.9 \\\hline
 458.sjeng	  & 400	        & 446	& 11.5& 0	& 0	& 0.0  & 127	& 135	  & 6.3 & 18	& 18	& 0.0  \\\hline
 462.libquantum	  & 362	        & 386	& 6.6 & 0 	& 0	& 0.0  & 87	& 87	  & 0.0 & 21	& 21	& 0.0  \\\hline
 464.h264ref	  & 17075	& 17153	& 0.5 & 4	& 4	& 0.0  & 338	& 382	  & 13.0& 72	& 72	& 0.0  \\\hline
 470.lbm	  & 39	        & 39	& 0.0 & 0	& 0	& 0.0  & 13	& 13	  & 0.0 & 4	& 4	& 0.0  \\\hline
 471.omnetpp	  & 565	        & 661	& 17.0& 32	& 30	& -6.3 & 4790	& 4843	  & 1.1 & 1321	& 1316	& -0.4 \\\hline
 473.astar	  & 691	        & 754	& 9.1 & 0	& 0	& 0.0  & 359	& 363	  & 1.1 & 113	& 113	& 0.0  \\\hline
 482.sphinx3	  & 1599	& 1693	& 5.9 & 11	& 9	& -18.2& 198	& 194	  & -2.0& 53	& 53	& 0.0  \\\hline
 483.xalancbmk	  & 9997	& 11453	& 14.6& 56	& 54	& -3.6 & 69534	& 68839	  & -1.0& 29875	& 29519	& -1.2 \\\hline
    \end{tabular}
  \end{center}
  \caption{Static metrics before (Base) and after global-code-motion (GCM) on SPEC2006 at -Ofast. The \% Column show percentage increase in metric w.r.t. Base. GCM improves LICM and inlining}
  \label{tab:stats}
\end{table}

Some useful statc metric are presented in Table~\ref{tab:stats}, which shows how code-motion
impacted important compiler optimizations like inlining and LICM. LICM removes loop invariant
code out of the loop. It may hoist the code to the loop's preheader or sink the code to loop's
post-dominator \cite{steven1997advanced}. As we can see number of instructions hoisted increased
significantly in several benchmarks, although number of instructions sunk did not change by much.
Number of inlined functiosn also improved for most benchmarks. In general when more functions
were inlined, more functions were deleted as in 400.perlbench, 445.gobmk.
